---
title: "Fars project"
author: "Filipe Duarte"
date: "2024-12-20"
output: rmarkdown::html_vignette
vignette: > 
  %\VignetteIndexEntry{Introduction to fars package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Carrega os pacotes 

```{r setup, include=FALSE}


knitr::opts_chunk$set(echo = TRUE)

#' Carreganto os pacotes necessários 

if(!require(dplyr)){install.packages("dplyr")}
if(!require(roxygen2)){install.packages("roxygen2")}
if(!require(tidyr)){install.packages("tidyr")}
if(!require(devtools)){install.packages("devtools")}
if(!require(readr)){install.packages("readr")}
if(!require(knitr)){install.packages("knitr")}
if(!require(maps)){install.packages("maps")}



```

 
#Introduction to fars

The fars package provides tools for analyzing and visualizing traffic accident data from the Fatality Analysis Reporting System (FARS). This system collects detailed information about traffic accidents across the United States, helping researchers and policymakers study trends, identify risks, and propose safety improvements.

Key Features
Load FARS data files.
Summarize traffic accidents by year and month.
Visualize accident patterns using maps.


##functions
fars_read
Reads a CSV file containing FARS data. If the file does not exist, an error is thrown.

```{r} 
 
#' Title : Read a fars_read data file
#'
#' Reads a CSV file containing fars data, a csv file and returns a tbl_df.
#' If the file does not exist, an error is thrown.
#'
#' @param filename A character string giving the name of the file to read, the file may exist or not.
#' @return A tibble containing the data from the CSV file. If the file does not exist an error message will be printed
#' @importFrom readr read_csv
#' @importFrom dplyr tbl_df
#' @export

fars_read <- function(filename) {
  if (!file.exists(filename))
    stop("file '", filename, "' does not exist")
  data <- suppressMessages({
    readr::read_csv(filename, progress = FALSE)
  })
  dplyr::tbl_df(data)
}


```


## make filename

Generates a filename for a specific year in the FARS dataset. 


```{r} 
#' Create a fars filename 
#'
#' Generates a filename for a specific year in the FARS dataset. This function permits to create
#' accident data files by year
#' @param year An integer or a value coercible to an integer representing the year.
#' @return A character string representing the filename.
#' @export


make_filename <- function(year) {
  year <- as.integer(year)
  sprintf("accident_%d.csv.bz2", year)
}

```


## fars_read_years

Reads FARS data files for multiple years and extracts the MONTH and year columns.

```{r} 

 
#' Read fars data for multiple years
#'
#' Reads FARS data files for multiple years and extracts the month and year columns.
#' If an invalid year is provided, a warning is issued, and the result for that year is `NULL`.
#'
#' @param years A vector of integers or values coercible to integers representing years.
#' @return A list of tibbles containing the `MONTH` and `year` columns for each valid year.
#' @importFrom dplyr mutate select
#' @export



fars_read_years <- function(years) {
  lapply(years, function(year) {
    file <- make_filename(year)
    tryCatch({
      dat <- fars_read(file)
      dplyr::mutate(dat, year = year) %>%
        dplyr::select(MONTH, year)
    }, error = function(e) {
      warning("invalid year: ", year)
      return(NULL)
    })
  })
}
```




## fars_summarize_years

Summarizes the number of accidents for each year and month.

```{r} 
 
#' Summarize fars data by year and month
#'
#' Summarizes the number of accidents for each year and month.
#'
#' @param years A vector of integers or values coercible to integers representing years.
#' @return A tibble summarizing the number of accidents for each month and year.
#' Columns represent years, and rows represent months.
#' @importFrom dplyr bind_rows group_by summarize
#' @importFrom tidyr spread
#' @export


fars_summarize_years <- function(years) {
  dat_list <- fars_read_years(years)
  dplyr::bind_rows(dat_list) %>%
    dplyr::group_by(year, MONTH) %>%
    dplyr::summarize(n = n()) %>%
    tidyr::spread(year, n)
}


```


## fars_map_state

Plots the locations of accidents in a specific state and year.

```{r}
  
#' Plot accidents on a state map
#'
#' Plots the locations of accidents in a specific state and year. 
#' If there are no accidents or an invalid state number is provided, an appropriate message is displayed.
#'
#' @param state.num An integer representing the state number.
#' @param year An integer or a value coercible to an integer representing the year.
#' @return A plot of accidents for the specified state and year. Returns `NULL` if no data is available.
#' @importFrom dplyr filter
#' @importFrom maps map
#' @importFrom graphics points
#' @export
#' @examples


fars_map_state <- function(state.num, year) {
  filename <- make_filename(year)
  data <- fars_read(filename)
  state.num <- as.integer(state.num)
  
  if (!(state.num %in% unique(data$STATE)))
    stop("invalid STATE number: ", state.num)
  data.sub <- dplyr::filter(data, STATE == state.num)
  if (nrow(data.sub) == 0L) {
    message("no accidents to plot")
    return(invisible(NULL))
  }
  is.na(data.sub$LONGITUD) <- data.sub$LONGITUD > 900
  is.na(data.sub$LATITUDE) <- data.sub$LATITUDE > 90
  with(data.sub, {
    maps::map("state", ylim = range(LATITUDE, na.rm = TRUE),
              xlim = range(LONGITUD, na.rm = TRUE))
    graphics::points(LONGITUD, LATITUDE, pch = 46)
  })
}


```



